Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-05 Thesis/02-chap2.qmd
fix chapter layout
current stage of last paper

```{r fig-heter-power-uniform-a, fig.height = 4, fig.cap = 'The effect of heteroskedasticity shape (parameter $a$) on the power of conventional and visual tests. The butterfly has higher power in both tests. Curiously, the visual test has a slightly higher power for the ``left-triangle" than the ``right-triangle" shape, when it would be expected that they should be similar, which is observed in conventional testing.'}

@@

# Limitations and Practicality

@@

# Conclusions

>>>>

```{r fig-heter-power-uniform-a, fig.height = 4, fig.cap = 'The effect of heteroskedasticity shape (parameter $a$) on the power of conventional and visual tests. The butterfly has higher power in both tests. Curiously, the visual test has a slightly higher power for the "left-triangle" than the "right-triangle" shape, when it would be expected that they should be similar, which is observed in conventional testing.'}

@@

## Limitations and Practicality

@@

## Conclusions

### 2024-08-05 Thesis/03-chap3.qmd
current stage of last paper

```{r fig-false-finding, fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretical "good" residual plot. This is further explained in @sec-model-distance-between-residual-plots. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

## Distance from a Theoretical "Good" Residual Plot  {#sec-model-distance-between-residual-plots}

@@

In this section, we introduce a distance measure between a true residual plot and a theoretical 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section @sec-model-distance-estimation.

@@

$$
\begin{aligned}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top(\boldsymbol{W})^{-1}\boldsymbol{\mu}_z\right),
\end{aligned}
$$ {#eq-kl-2}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})$. The assumed error variance $\sigma^2$ is set to be $tr(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

@@

To evaluate the Kullback-Leibler divergence of non-normal $P$ from $Q$, the fallback is to solve @eq-kl-1 numerically. However, since $\boldsymbol{e}$ is a linear transformation of non-normal random variables, it is very common that the general form of $P$ is unknown, meaning that we can not easily compute $p(\boldsymbol{e})$ using a well-known probability density function. Additionally, even if $p(\boldsymbol{e})$ can be calculated for any $\boldsymbol{e} \in \mathbb{R}^n$, it will be very difficult to do numerical integration over the $n$ dimensional space, because $n$ could be potentially very large.

@@

Given $\boldsymbol{X}$ and $\boldsymbol{\beta}$, the algorithm for approximating @eq-kl-1 starts from simulating $m$ sets of observed values $\boldsymbol{y}$ according to the data generating process. The observed values are stored in a matrix $\boldsymbol{A}$ with $n$ rows and $m$ columns, where each column of $\boldsymbol{A}$ is a set of observed values. Then, we can get $m$ sets of realized values of $\boldsymbol{e}$ stored in the matrix $\boldsymbol{B}$ by applying the residual operator $\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}$. Furthermore, kernel density estimation (KDE) with Gaussian kernel and optimal bandwidth selected by the Silverman's rule of thumb [@silverman2018density] is applied on each row of $B$ to estimate $p_i(e_i)$ for $i = 1, ..., n$. The KDE computation can be done by the `density` function in R.

@@

\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $B_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $B$, $\hat{p_i}(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\widehat{\sigma^2} = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.

@@

In the previous sections, we have defined a distance measure given in @eq-kl-0 for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. It can be noticed that this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

@@

With the estimated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a theoretical "good" residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

@@

Note that @eq-kl-0 might be influenced by the number of random variables involved in its evaluation. Generally, a larger number of observations will lead to a greater distance $D$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

For a consistent data generating process, $D$ typically increases logarithmically with the number of observations. This behaviour comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

@@

```{r fig-delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs $\\delta$-differnence plot. The brown line is smoothing curve produced by fitting gnealized additive models", fig.height = 4}
experiment %>%
  left_join(lineup_vss %>%
              group_by(unique_lineup_id) %>%
              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
  ggplot() +
  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
  ylab("Weighted detection rate") +
  xlab(expression(delta-difference)) +
  theme_light()
```

>>>>

```{r fig-false-finding, fig.width = 4, fig.height = 4*4/5, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0 corresponds to the x-intercept, i.e. $y=0$). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity, however, here the result is due to skewed distribution of the predictors rather than heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

In this study, we chose to define and use a distance between a true residual plot and a theoretically "good" residual plot. This is further explained in @sec-model-distance-between-residual-plots. @vo2016localizing have also demonstrated that defining a proper distance between images can enhance the matching accuracy in image search compared to a binary outcome model.

## Distance from a Theoretically "Good" Residual Plot  {#sec-model-distance-between-residual-plots}

@@

In this section, we introduce a distance measure between a true residual plot and a theoretically 'good' residual plot. This measure quantifies the divergence between the residual distribution of a given fitted regression model and that of a correctly specified model. The computation assumes knowledge of the data generating processes for predictors and response variables. Since these processes are often unknown in practice, we will discuss a method to estimate this distance using a computer vision model in Section @sec-model-distance-estimation.

@@

$$
\begin{aligned}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\boldsymbol{W}|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\boldsymbol{W}^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z^\top\boldsymbol{W}^{-1}\boldsymbol{\mu}_z\right),
\end{aligned}
$$ {#eq-kl-2}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R})$. The assumed error variance $\sigma^2$ is set to be $\text{tr}(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

@@

To evaluate the Kullback-Leibler divergence of non-normal $P$ from $Q$, the fallback is to solve @eq-kl-1 numerically. However, since $\boldsymbol{e}$ is a linear transformation of non-normal random variables, it is very common that the general form of $P$ is unknown, meaning that we can not easily compute $p(\boldsymbol{e})$ using a well-known probability density function. Additionally, even if $p(\boldsymbol{e})$ can be calculated for any $\boldsymbol{e} \in \mathbb{R}^n$, it will be very difficult to do numerical integration over the $n$-dimensional space, because $n$ could be potentially very large.

@@

Given $\boldsymbol{X}$ and $\boldsymbol{\beta}$, the algorithm for approximating @eq-kl-1 starts from simulating $m$ sets of observed values $\boldsymbol{y}$ according to the data generating process. The observed values are stored in a matrix $\boldsymbol{A}$ with $n$ rows and $m$ columns, where each column of $\boldsymbol{A}$ is a set of observed values. Then, we can get $m$ sets of realized values of $\boldsymbol{e}$ stored in the matrix $\boldsymbol{B}$ by applying the residual operator $\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}$. Furthermore, kernel density estimation (KDE) with Gaussian kernel and optimal bandwidth selected by the Silverman's rule of thumb [@silverman2018density] is applied on each row of $\boldsymbol{B}$ to estimate $p_i(e_i)$ for $i = 1, ..., n$. The KDE computation can be done by the `density` function in R.

@@

\noindent where $\hat{D}_{KL}^{(i)}$ is the estimator of the Kullback-Leibler divergence for an individual residual $e_i$, $\boldsymbol{B}_{ij}$ is the $i$-th row and $j$-th column entry of the matrix $\boldsymbol{B}$, $\hat{p_i}(.)$ is the kernel density estimator of $p_i(.)$, $q(.)$ is the normal density function with mean zero and an assumed variance estimated as $\widehat{\sigma^2} = \sum_{b \in vec(\boldsymbol{B})}(b - \sum_{b \in vec(\boldsymbol{B})} b/nm)^2/(nm - 1)$, and $vec(.)$ is the vectorization operator which turns a $n \times m$ matrix into a $nm \times 1$ column vector by stacking the columns of the matrix on top of each other.

@@

In the previous sections, we have defined a distance measure given in @eq-kl-0 for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. However, this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

@@

With the estimated distance $\hat{D}$, we can compare the underlying distribution of the residuals to a theoretically "good" residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

@@

Note that if more observations are used for estimating the linear regression, the result of @eq-kl-1 will increase, as the integration will be performed over a higher-dimensional space. For a consistent data generating process, $D$ typically increases logarithmically with the number of observations. This behaviour comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

Since $\hat{D}$ is an estimate of $D$, it is expected that a larger number of observations will also lead to a higher $\hat{D}$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

@@

```{r fig-delta, fig.pos = "!h", fig.cap = "A weighted detection rate vs $\\delta$-difference plot. The brown line is smoothing curve produced by fitting gnealized additive models", fig.height = 4}
experiment %>%
  left_join(lineup_vss %>%
              group_by(unique_lineup_id) %>%
              summarise(delta_diff = first(delta_diff), gamma_diff = first(gamma_diff))) %>%
  ggplot() +
  geom_vline(xintercept = 0, alpha = 0.4, linetype = 2) +
  geom_point(aes(delta_diff, prop_detect), alpha = 0.3) +
  geom_smooth(aes(delta_diff, prop_detect), se = FALSE, col = "#994F00") +
  ylab("Weighted detection rate") +
  xlab(expression(delta-difference)) +
  theme_light()
```
